# Vectorization
Most of R's functions are vectorized.

 - These functions will operate on all elements of a vector at once
  - No need to loop through and act on each element one at a time
 
**Advantages of vectorisation**
We can write code that is:

  - concise 
  - easy to read and
  - less error prone.

e.g.
```{r vectorisation-multiplication}
x <- 1:4
x
x * 2   
```

Above, each element of the vector was multiplied by 2.  
Lets see how vectorisation works while adding two vectors in `R`.


```{r vectorisation-add-two-vectors}
y <- 6:9
y
x + y
```
Here, each element of `x` was added to its corresponding 
element of `y`. Lets see that again:

```{r}
x; y; x+y

```

Another way to implement vectorisation while adding two vectors together is by using a `for` loop.


## Structure of a `for()` loop
```{r eval=FALSE}
for (iterator in set_of_values) {
         do a thing
     } 
```
 - indentation is not important in R compared to Shell and Python
 - but its a good practice to use indentation

**Uses of for() loop**

 - repeating operations
 
 
**Example of a for loop** 
```{r for-loop}
for (i in 1:10) {
    print(i)
}
```


`for()` loop is good for iterating over a set of values, when the order of iteration is important. We saw `for()` loop yesterday.

Its good to learn for() loops, but avoid using for() loops unless the order of iteration is important. 

Here's an example of a for loop that does vector addition.

```{r vect-addition-for-loop}
x
y
#vectorisation using for loop
output_vector <- c()
for (i in 1:4) {
    output_vector[i] <- x[i] + y[i]
}
output_vector
```


Compare this to the output using vectorised operations.
```{r}
x
y
sum_xy <- x + y
sum_xy

identical(output_vector, sum_xy)
```

Both are identical.


Let's try Vectorisation in gapminder data. 
Let's load the gapminder data that we used yesterday in the first part of our R lesson. 



You can clear your environment for a fresh start by removing all the objects using the below command.

```{r}
rm(list = ls())
```
Please import the `gapminder` data to your `RStudio` environment. If you do not have the `gapminder` loaded into your RStudio session, you can do it using the below code chunk. This code chunk is also provided in the [collaborative document](https://pad.carpentries.org/2020-07-22-usda-online).


```{r loading-gapminder-data}
# loading data using a link
gapminder <- read.csv("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/gapminder_data.csv", stringsAsFactors=FALSE)

# if the gapminder is already in your data folder
# you can load it using the below code:
#gapminder <- read.csv("data/gapminder_data.csv", stringsAsFactors=FALSE)
```



Examine the gapminder dataset.

```{r}
head(gapminder)  # shows the first few rows of the dataset
tail(gapminder)  # shows the last few rows of the dataset
```

Lets  try vectorisation on the `pop` column of the `gapminder` dataset.

 1. Make a new column in the `gapminder` data frame that
 contains population in units of millions of people.

 2. Check the head or tail of the data frame to make sure
 it worked.

```{r}

gapminder$pop_millions <- gapminder["pop"] / 1e6
#or
gapminder$pop_millions_2 <- gapminder$pop / 1e6
# or
gapminder["pop_millions_3"] <- gapminder$pop / 1e6
```

Check the head or tail of the data frame to make sure
it worked.
```{r}
head(gapminder)
```


```{r}
# check if both the created variables are identical or not
# using the R function identical()
identical(gapminder$pop_millions, gapminder$pop_millions_2 )
```


In addition to mathematical operators, Comparison operators, logical operators & many functions are also vectorized:


Let's see the **Comparison operators**

```{r}
x <- 0:6
x
x > 2 
```

**Logical operators**
```{r}
a <- x > 3  # or, for clarity, a <- (x > 3)
a
```



**Functions**
`log()`
```{r}
x
x <- 1:4
x
log(x)
```



**Vectorisation in matrices**
Vectorized operations work element-wise on matrices.

```{r}
m <- matrix(1:12, nrow=3, ncol=4)
m
m * -1 # element wise operation

m
m ^ -1 # element wise operation
```



**Very important:**

  - the operator `*` = > element-wise multiplication
  - for matrix multiplication, use the `%*%` operator
```{r}
m
matrix(1, nrow=4, ncol=1)

m %*% matrix(1, nrow=4, ncol=1)

matrix(1:4, nrow=1) #row matrix
matrix(1:4, nrow=1) %*% matrix(1:4, ncol=1)
```

