# Functions

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


**Functions**

 - are the basic building block of most programming languages
 - are used to gather a sequence of operations into a whole, preserving it for ongoing use.
 - have names that we can remember and call 
 - have a defined set of inputs and expected outputs 
 - provide connections to the larger programming environment


We can repeat several operations with a single function call.

There are 2 types of functions:

 1. built-in functions
 2. user-defined functions


Below are some of the built-in functions from **base R**.

```{r, eval=FALSE}
log()
mean()
min()
sd()
```

 



User-defined functions are created in **R** using the keyword __function__. See below regarding how to get help for writing functions in **R**.



```{r keyword-function}
?"function"
```



## Components of a function
Below presented are the components of a function:

a. name
b. arguments/inputs - within parentheses -()
c. body - within curly braces - {}
      - statements that are executed when it runs

The statements in the body are indented by two spaces as a good practice for writing `R` code. A function is called in `R` by providing the function name with the arguments in parenthesis, like shown below.


```{r eval=FALSE}
# defining a function
function_name <- function(arguments){
    body
}

# this is how we call or use a function
function_name(arguments)
```


Let’s define a function `fahr_to_kelvin()` that converts 
temperatures from Fahrenheit to Kelvin:


```{r, fahr_to_kelvin}
fahr_to_kelvin <- function(temperature) {
    kelvin <- ((temperature - 32) * (5 / 9)) + 273.15
    return(kelvin)
}
```

Run the above code chunk on your RStudio console. 

Next, check if you have executed the code chunk by going through the `History` tab of the top right panel on your RStudio window. 
You can also make sure that the function named `fahr_to_kelvin` is available for you to use by searching the **RStudio** `Environment`. From the top right panel of your RStudio window, go the tab `Environment` and search for the function `fahr_to_kelvin` under the Function names.

Lets try using our function to convert the freezing point of water from Fahrenheit to kelvin. Freezing point of water is 0 °C or 32 °F.


```{r fahr_to_kelvin-1}
fahr_to_kelvin(32)
```


Next, use the same function to convert the boiling point of
water from Fahrenheit to Kelvin. Note that the boiling point of water is 100 °C (212 °F).

```{r fahr_to_kelvin_bp}
fahr_to_kelvin(212)
```



## Challenge 1
Write a function called kelvin_to_celsius() that takes a temperature in Kelvin & converts that to temperature in Celsius.

To convert from Kelvin to Celsius you subtract 273.15

Reference: http://swcarpentry.github.io/r-novice-gapminder/10-functions/index.html

```{r challenge-1, eval=FALSE}
kelvin_to_celsius <- function(temp) {
    .....
    return(celsius)
}
```



**Solution**
```{r challenge-1-solution}
kelvin_to_celsius <- function(temp) {
    celsius <- temp - 273.15
    return(celsius)
}
```


## Combining functions

We have already defined two functions:

1. `fahr_to_kelvin()`  - converts from Fahrenheit to Kelvin
2. `kelvin_to_celsius()` - converts from Kelvin to Celsius

See if they are available in your global environment of RStudio.


The real power of functions comes from mixing, 
matching and combining them into ever-larger chunks
to get the effect we want.


Lets try to understand how to combine functions.

We will define a third function to convert directly from 
Fahrenheit to Celsius, by reusing the two functions above.
Let the name of third function be `fahr_to_celsius()`

- input: temperature in Fahrenheit
- expected output: temperature in Celsius


We are going to accomplish this in 3 steps:

1. Convert Fahrenheit to Kelvin : `fahr_to_kelvin()`
2. Convert Kelvin from Step1 to Celsius: `kelvin_to_celsius()`
3. Return the above result at the end of the function

```{r combining-functions}
fahr_to_celsius <- function(temp) {
    temp_k <- fahr_to_kelvin(temp)
    result <- kelvin_to_celsius(temp_k)
    return(result)
}

# or can nest functions like this
fahr_to_celsius_2 <- function(temp) {
    temp_c <- kelvin_to_celsius(fahr_to_kelvin(temp))
    return(temp_c)
}
```

Lets try to convert the boiling and freezing point of
water from Fahrenheit to Celsius. First, make sure all the three functions are available in your global environment.

Call your combined function for the above mentioned conversions as shown below 
```{r using-combined-function}
fahr_to_celsius(212) # boiling point of water
fahr_to_celsius(32) # freezing point of water
```





## Defensive Programming
- Checking function parameters
-  writing functions provides an efficient way to make R code
  re-usable and modular

-  it is important to ensure that functions only work in their
    intended use-cases. 

   - frequently check function parameters using conditions
     and throw an error if something is wrong. 
  
   - These checks are referred to as assertion statements because 
     we want to assert some condition is TRUE before proceeding. 

  -  make it easier to debug because they give us a better idea of 
     where the errors originate.

 - How to use assertion in `R` ?
       - we can tell `R` to stop executing the function using stop()




### Case 1: Functions with one argument
We can create conditions with `stopifnot()`. Let’s start by re-examining `fahr_to_kelvin()`.

- Here, we know that the argument `temp` must be a `numeric`
      - Otherwise, the mathematical operations used in the temperature conversions will not work. 
- Hence, we check this with an `if` statement and `stop()` 
if the condition is violated

```{r testing-is-numeric}
is.numeric("TRUE")
is.numeric(TRUE)
is.numeric(2)
```


```{r defensive-fahr_to_kelvin}
fahr_to_kelvin <- function(temp) {
    if (!is.numeric(temp)) {
        stop("temp must be a numeric vector.")
    }
    kelvin <- ((temp - 32) * (5 / 9)) + 273.15
    return(kelvin)
}
```

### Case 2: Multiple arguments or conditions
  - Use `stopifnot()` for type-assertion
  - works in the same way as `stop()`
  - shorter syntax
  - throws an error if it finds a condition that is `FALSE`. 

Here is the code for defensive programming with `stopifnot()`

```{r}
fahr_to_kelvin_stopifnot <- function(temp) {
    stopifnot(is.numeric(temp))
    kelvin <- ((temp - 32) * (5 / 9)) + 273.15
    return(kelvin)
}
```

**Note**
  - We can list as many conditions that should evaluate to `TRUE`
  - Secondary purpose: extra documentation for the function.

Lets try checking the impact of the newly added assertions
on our function `fahr_to_kelvin()`.

- Check the Global environment

If you type your function name without parenthesis, R will print out the code for your function.
```{r }
fahr_to_kelvin_stopifnot
```



**Freezing point of water**

```{r }
fahr_to_kelvin(temp = 32)
fahr_to_kelvin_stopifnot(temp = 32)
```


Let's convert the number `32` to the type `factor` using the function `as.factor()`. This is an example of a type coercion that we saw on day 1 of the R lesson.
```{r}
as.factor(32)
```


The metric is a factor instead of numeric

```{r eval=FALSE}
fahr_to_kelvin(temp = as.factor(32))
fahr_to_kelvin_stopifnot(temp = as.factor(32))
```
Both functions still works when given proper input.
But fails instantly if given improper input.




We will define a function `calcGDP()` that calculates the Gross Domestic Product of a nation from the `gapminder` dataset that we used in the chapter `Vectorisation`.

`calcGDP` takes a dataset and multiplies the population column
`pop` with the GDP per capita column `gdpPercap` to get the GDP.


```{r calcGDP}
head(gapminder)

calcGDP <- function(dat) {
    gdp <- dat$pop * dat$gdpPercap
    return(gdp)
}
```



Lets see what this function returns when applied to a part of the gapminder data.
```{r}
calcGDP(head(gapminder))
```


As we want our results to be more informative than the above, let's add some more arguments so we can extract per year and country GDP. We can also apply 
1. defensive programming &
2. default values for the arguments

```{r, calcGDP-defensive-default}
calcGDP <- function(dat, year=NULL, country=NULL) {
    if(!is.null(year)) {
        dat <- dat[dat$year %in% year, ]
    }
    if (!is.null(country)) {
        dat <- dat[dat$country %in% country,]
    }
    gdp <- dat$pop * dat$gdpPercap
    
    new <- cbind(dat, gdp=gdp)
    # modified gapminder with a gdp column
    return(new)
}
```
Explanation of what the function `calcGDP` does:

i) subsets the provided data by year if the year argument isn’t empty
ii) subsets the result by country if the country argument isn’t empty. 
iii) calculates the GDP for the resultant subset
iv) adds the GDP as a new column to the subsetted data
v) returns the subsetted data (if year or/and country was present in the input dataframe) with a gdp column added as the final result. 


**Note**: Implementing defensive programming can make our functions more flexible for later use. 


We can use it to calculate the GDP for:
     
- The whole dataset;
- A single year;
- A single country;
- A single combination of year and country.

By using %in% instead, we can also give multiple years or countries to those arguments. Please see help for `%in%` which we covered yesterday using the below functions:

```{r}
help("%in%")
?"%in%"
```




Let’s take a look at what happens when we specify the year:

```{r}
head(calcGDP(gapminder, year = 2007))
```

Or for a specific country:
```{r}
calcGDP(gapminder, country="Australia")
```






Let's do Challenges 4 & 5

Reference: http://swcarpentry.github.io/r-novice-gapminder/10-functions/index.html


## Challenge 4

Test out your GDP function by calculating the GDP for New Zealand in 1987. How does this differ from New Zealand’s GDP in 1952?

```{r solution-frame-work, eval=FALSE}
calcGDP(gapminder, year = ..., country = ..)
```



```{r}
# see if you get any results with these options. If you don't get results, can you tell why ?
calcGDP(gapminder, year = 1952, country = "Newzealand")
calcGDP(gapminder, year = 1952, country = "NewZealand")
calcGDP(gapminder, year = 1952, country = "New Zealand")
```




## Challenge 5

Write a function called fence() that takes two vectors as arguments, called text and wrapper, and prints out the text wrapped with the wrapper:

The paste() function, we learned yesterday, can be used to combine text together. See below.

```{r paste-function-demo}
best_practice <- c("Write", "programs", "for", "people", "not", "computers")
paste(best_practice, collapse=" ")
```

Let the argument `text` be `best_practice` and `wrapper` be `****`. The frame for the function is given below. You can try completing this.


```{r, eval=FALSE }
fence <- function(text, wrapper){
    ...
    ....
    return(result)
}
# calling the function
fence(text=.., wrapper=..)
```






**Solution**


```{r, challenge-5-solution}
fence <- function(text, wrapper){
    text <- c(wrapper, text, wrapper)
    result <- paste(text, collapse = " ")
    return(result)
}

best_practice <- c("Write", "programs", "for", "people", "not", "computers")
fence(text=best_practice, wrapper="***")
```



## How to use functions for later analyses

1. Save functions to a file: e.g.: "functions.R"
2. Use it in another R script by using the `source()` function
    - `source("~/functions/functions.R")`


## Other important concepts
- Pass by value
- Function scope
- Testing and documenting



    



