# Saving plots and Writing data

## Saving plots in RStudio

i. **`ggsave()`**
```{r}
library(ggplot2)
```


```{r eval=FALSE}
ggplot(data=gapminder,
       aes(x=year, y=lifeExp, colour=country)) +
    geom_line() +
    theme(legend.position = "none")
ggsave("PlotName.pdf")
```

ii. **Export from RStudio**
  - Make your plot in the RStudio
  - Go to the lower right panel 
  - Select the tab `Plots`
  - Select the options for export from the drop-down menu
    - Save as image or Save as pdf  or Copy to clipboard

iii. **Plotting devices**

Plotting devices can be used in the `R` and `RStudio` to save plots. We can control the size and resolution of the plots to be saved using arguments to the plotting devices. One such plotting device is `pdf`. The below code chunk can be used to save the `ggplot` as a `pdf` file with the specified height and width.



```{r eval=FALSE}
pdf("Life_Exp_vs_time.pdf", width=12, height=4)
ggplot(data=gapminder,
       aes(x=year, y=lifeExp, colour=country)) +
    geom_line() +
    theme(legend.position = "none")

# Make sure to turn off the pdf device.
# Else, the plot won't be saved.
dev.off()
```

`jpeg`, `png`, `tiff` etc. are other plotting devices available in `R` to save plots. They can be called with arguments similar to the `pdf()`. 

```{r eval=FALSE}
jpeg(..)
png(..)
tiff(..)
```

## Writing Table

We can use the `write.table()` function for writing tables, similar to `read.table()` we used for importing data.

We can select the gapminder data for Australia and write it to a table.

```{r subset-and-write-table}
# create a directory named cleaned-data if its not there
if(!dir.exists("./cleaned-data")){
  dir.create("cleaned-data")
}
#subset the data
aust_subset <- gapminder[gapminder$country == "Australia",]
#write the data
write.table(aust_subset,
    file="cleaned-data/gapminder-aus.csv",
    sep=","
)
```
Lets use the **`Terminal`** tab on the lower left panel of the RStudio and check
the data.

```{r eval=FALSE}
head cleaned-data/gapminder-aus.csv
```
For help on the `write.table` function, use the `help` function.

```{r}
help(write.table)
```

Lets use additional arguments for the `write.table` function so that quotes are not present for the values (option `quotes=FALSE`) and row names are not written to the cleaned file.

```{r}
write.table(
    gapminder[gapminder$country == "Australia",],
    file="cleaned-data/gapminder-aus.csv",
    sep=",", quote=FALSE, row.names=FALSE
)
```

Lets use the RStudio **`Terminal`** again and examine the cleaned file.
```{r, eval=FALSE}
# note - this command is for using on the Terminal window
# and not on the RStudio console
head cleaned-data/gapminder-aus.csv
```


