--- 
title: "SWC R Workshop @ USDA: Notes and Code"
author: "Preethy Nair"
date: "July 23, 2020"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: rstudio/bookdown-demo
description: "This is a supplementary note for the SWC worshop at the USDA. The output format is bookdown::gitbook."
---

# Information

This is a supplementary notebook for the online SWC R workshop at the USDA on July 23, 2020. The notebook includes the code that was used during workshop and notes based on the Carpentries R lesson, [R for Reproducbile data analyis](https://swcarpentry.github.io/r-novice-gapminder/).

Please see the [course website](https://nairps.github.io/2020-07-22-usda-online/) to access other lessons we covered for the workshop. The workshop notes for other sessions are provided in the [collaborative document](https://pad.carpentries.org/2020-07-22-usda-online).  


Summary of yesterday's `R` lessons are also provided in the same [collaborative note](https://pad.carpentries.org/2020-07-22-usda-online).

To get help for a function in `R`, we can use `help`.

```{r eval=FALSE}
help("%in%") # value matching
help("c")  # combine function
help("%>%") # pipe operator
help("<-") # assignment operator
help("-") # arithmetic operator
```


## Prerequisites
To participate in this workshop session on **R**, you need to install **R**, **RStudio** and some packages from `R` [@R-base] (`dplyr` [@R-dplyr], `tidyr`[@R-tidyr], `knitr` [@R-knitr; @xie2015]).
Please see the [course website](https://nairps.github.io/2020-07-22-usda-online/) for instructions to install `R` and `RStudio`. 
Once you have **R** installed on your computers, the packages required for this session can be installed using the below commands.

```{r eval=FALSE}
install.packages(c("dplyr", "tidyr", "knitr"))
# or individually
install.packages("dplyr")
install.packages("tidyr")
install.packages("knitr")
```

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'dplyr', 'tidyr'
), 'packages.bib')
```
